---
title: "A4-vat-tconv-treg"
author: "JRD"
date: "February 9, 2016"
output: html_document
---

last modified: `r Sys.Date()`
<br>    

#### R Markdown Specifications
Set global options
```{r setoptions, echo=F}
library(knitr)
opts_chunk$set(echo=F, eval=F, warning=F, message=F)
```
<br>    


Github
https://github.com/jdis24/ATAC.git
<br>    

#### Pseudo
+ [these steps were done in "ATAC_peaks.Rmd"]
+ for each sample, call peaks in each replicate
+ then merge into single peakset
+ for each replicate, count reads in each peak, normalized to total library size
+ export .csv files of peaks and counts

+ [these steps done in this document]
+ plot distributions of read counts 
+ use DiffBind to find differential peaks between VAT Treg and Tconv
+ use merged peaks as input to DiffBind, with .bam files for each replicate
+ export all peaks in the contrast with p-values and/or FDR
+ plot p-value/FDR distributions
+ set cutoffs
+ export differential and similar peaks, with counts in rpm
<br>    


#### Import peak files into peak list
```{r peak_list, eval=T}
# create path to .csv files  
path <- "/Volumes/JD_WD2/JRD_CBDMLab/1_JRD_CBDM_PROJECTS/ProjectFolders/T-TregEpigenome/Data/1_Experimental/ATAC/Master_peaks/R/peaks_with_counts/working/"

# get filenames
filenames1 <- list.files(path = path, pattern = ".csv")
filenames <- paste(path, filenames1, sep="") # update filenames with path

# combine peak table .csv dfs into a list
colClass <- c(rep("factor",2), rep("numeric",3), "character", rep("numeric",4)) # set column classes
pl <- lapply(filenames, read.csv, colClasses = colClass) # [p]eak [l]ist

# name the list elements to match the files
names(pl) <- gsub("\\.csv", "", filenames1)
```
<br>    


#### Generate table of peak numbers for range of rpm threshold values
```{r threshold_table, eval=T, results="asis"}
library(xtable)

# function to compute the number of peaks >=n rpm in each sample
# ARGS: df = dataframe of peak height rpms, peak_height_var = string of var name storing rpm values, cutoffs = numeric vector of rpm cutoffs 
# USAGE: number_peaks_above_nrpm(df, "peak_height_var", cutoffs)
number_peaks_above_nrpm <- function(df, rpm_var, cutoffs) { 
  # create indexing vector representing number of elements in cutoffs
  n <- length(cutoffs) 
  # create output vector whose length = length of cutoffs
  o <- numeric(n) 
  # create variable to store vector of extracted rpm values
  rpm <- df[[rpm_var]]
  # loop over each element of cutoffs and count number rows where peak height >= that element
  for(i in 1:n) { 
    o[i] <- as.numeric(sum(rpm >= cutoffs[[i]])) # calculate the total number of rows where rpm >= ith value in cutoffs  
  }
  # print output vector
  return(o) 
}


# create cutoff vector of rpm values
cutoffs <- c(seq(0,5,by=0.5))

# apply cutoffs to list
peaks_pass_filter <- sapply(pl, number_peaks_above_nrpm, "rpm", cutoffs)

# transform list into dataframe (if doesn't work may have to call base as.data.frame bc of Biocgenerics masking?)
ppf <- as.data.frame(peaks_pass_filter) #[p]eaks [p]ass [f]ilter
ppf <- cbind(ppf,"rpm_cutoff" = cutoffs) # add column for each cutoff

# write to file
write.csv(ppf, "peak_totals_rpm_filtered.csv", row.names=F)

# generate xtable
print(xtable(ppf), type="html", include.rownames=F)
```
<br>    


#### Plot distribution of read counts for each sample
```{r plots, eval=T}
library(reshape2)
library(ggplot2)
library(dplyr)
# set path for plots 
path <- "/Volumes/JD_WD2/JRD_CBDMLab/1_JRD_CBDM_PROJECTS/ProjectFolders/T-TregEpigenome/Data/1_Experimental/ATAC/Analyses/A4-vat-tconv-treg/"

# get peak list into shape for plotting
plm <- melt(pl, id.vars=names(pl[[1]])) # [p]eak [l]ist [m]elted
plm$L1 <- NULL # remove extraneous column

# calculcate log2 rpm values and append as new column
plm$log2_rpm <- log(plm$rpm, 2)

# show number of peaks where rpm =0 
rpm0pks <- plm$log2_rpm == "-Inf"
plm_rpm0pks <- plm[rpm0pks,]

print("Peaks where rpm =0 ...") 
xtable(plm_rpm0pks)

# filter out peaks where rpm = 0
plmf <- plm[!rpm0pks,]  #plm[f]iltered

# draw boxplot of all peaks in each sample
g <- ggplot(data = plmf, aes(x=Sample, y=log2_rpm, fill=Sample))
g + geom_boxplot(varwidth =T) +
  theme_bw() + 
  ylim(-2,10) +
  theme(axis.title.x = element_blank())
ggsave(paste0(path, "a4-boxplot.pdf"), width=5, height=4) # export pdf

# draw boxplot of peaks passing n rpm
# define function to filter peaks and draw plot
bp_nrpm <- function(df, n) {
  df %>%
    filter(rpm >= n) %>% # filter by rpm cutoff 
    ggplot(aes(x=Sample, y=log2_rpm, fill=Sample)) + # plot
    geom_boxplot(varwidth=T) +
    ylim(-2,10) +
    theme_bw() + 
    theme(axis.title.x = element_blank())
}

n <- 3 # set rpm cutoff
bp_nrpm(plmf, n) # draw plot
ggsave(paste0(path,"a4-boxplot-",n,"-rpm-filtered.pdf"), width =5, height =4) # export pdf
# ggsave(paste0(path, "a4-boxplot-rpm-filtered.pdf"), width=5, height=4)


# draw histograms of all peaks in each sample
g <- ggplot(data = plmf, aes(x=log2_rpm))
g + geom_histogram(fill="white",color="black",binwidth=0.1) +
  facet_grid(Sample~.) +
  geom_vline(xintercept =log(1,2)) +
  geom_vline(xintercept = log(2,2)) +
  geom_vline(xintercept = log(3,2)) +
  theme_bw()
ggsave(paste0(path, "a4-count-hist.pdf"), width=5, height=6.5)
```





