---
title: "A7-all-ttreg"
author: "JRD"
date: "March 7, 2016"
output: html_document
---

last modified: `r Sys.Date()`
<br>    

#### README
+ this document contains code and explanation for an analysis comparing Colon, Muscle, Splenic and VAT Treg
+ for details on how the input files used in this analysis were generated, see *ATAC_mapping.Rmd* and *ATAC_peaks.Rmd*
+ **difference between A5 and A6 is latter filters peaks by both min signal and replicate CV**
<br>    


#### R Markdown Specifications
Set global options
```{r setoptions, echo=F}
library(knitr)
opts_chunk$set(echo=F, eval=F, warning=F, message=F)
```
<br>    

Github
https://github.com/jdis24/ATAC.git
<br>    


#### Pseudo
**call peaks and get read counts**
+ [these steps were done in "ATAC_peaks.Rmd"]
+ for each sample, call peaks in each replicate
+ then merge into single peakset
+ for each replicate, count reads in each peak, normalized to total library size
+ export .csv files of peaks and counts
**filter for strong peaks with low variance, merge peaks and get read counts for merged peaks**
+ [all subsequent steps done in this document]
+ plot distributions of read counts 
+ plot **mean vs replicate cv** of read counts for each sample 
+ filter for strong peaks (start with >=2.5 rpm) with low cv (?)
+ **check that filters worked on genome browser**
+ get genomic annotations of confident strong peaks (compare to unfilterd to check degree of TSS enrichment)
+ use DiffBind to create merged peakset across all T-treg and get read counts
+ export all peaks and their rpkm score in each sample 
<br>  


#### Import peak files into peak list and export bed files
```{r peak_list, eval=T}
# create path to .csv files  
path <- "/Volumes/JD_WD2/JRD_CBDMLab/1_JRD_CBDM_PROJECTS/ProjectFolders/T-TregEpigenome/Data/1_Experimental/ATAC/Master_peaks/R/peaks_with_counts/working/"

# get filenames
filenames1 <- list.files(path = path, pattern = ".csv")
filenames <- paste(path, filenames1, sep="") # update filenames with path

# combine peak table .csv dfs into a list
colClass <- c(rep("factor",2), rep("numeric",3), "character", rep("numeric",4)) # set column classes
pl <- lapply(filenames, read.csv, colClasses = colClass) # [p]eak [l]ist

# name the list elements to match the files
names(pl) <- gsub("\\.csv", "", filenames1)

# convert each df to bed format
pl_bed <- lapply(pl, function(x) {
  x <- x[,c("seqnames","start","end")]
})

# export each df as tdt bed file
path <- "/Volumes/JD_WD2/JRD_CBDMLab/1_JRD_CBDM_PROJECTS/ProjectFolders/T-TregEpigenome/Data/1_Experimental/ATAC/Analyses/A7-all-ttreg/bed/"

lapply(names(pl_bed),
       function(x) write.table(pl_bed[x], file=paste(paste0(path,x),"bed",sep="."), sep = "\t", 
                              quote=F, row.names=F,col.names=c("#chr","start","end")))
```
<br>    


#### Generate table of peak numbers for range of rpm threshold values
```{r threshold_table, eval=T, results="asis"}
library(xtable)

# function to compute the number of peaks >=n rpm in each sample
# ARGS: df = dataframe of peak height rpms, peak_height_var = string of var name storing rpm values, cutoffs = numeric vector of rpm cutoffs 
# USAGE: number_peaks_above_nrpm(df, "peak_height_var", cutoffs)
number_peaks_above_nrpm <- function(df, rpm_var, cutoffs) { 
  # create indexing vector representing number of elements in cutoffs
  n <- length(cutoffs) 
  # create output vector whose length = length of cutoffs
  o <- numeric(n) 
  # create variable to store vector of extracted rpm values
  rpm <- df[[rpm_var]]
  # loop over each element of cutoffs and count number rows where peak height >= that element
  for(i in 1:n) { 
    o[i] <- as.numeric(sum(rpm >= cutoffs[[i]])) # calculate the total number of rows where rpm >= ith value in cutoffs  
  }
  # print output vector
  return(o) 
}


# create cutoff vector of rpm values
cutoffs <- c(seq(0,5,by=0.5))

# apply cutoffs to list
peaks_pass_filter <- sapply(pl, number_peaks_above_nrpm, "rpm", cutoffs)

# transform list into dataframe 
ppf <- as.data.frame(peaks_pass_filter) #[p]eaks [p]ass [f]ilter
ppf <- cbind(ppf,"rpm_cutoff" = cutoffs) # add column for each cutoff

# write to file
path <- "/Volumes/JD_WD2/JRD_CBDMLab/1_JRD_CBDM_PROJECTS/ProjectFolders/T-TregEpigenome/Data/1_Experimental/ATAC/Analyses/A7-all-ttreg/"
write.csv(ppf, file = paste0(path,"peak_totals_rpm_filtered.csv"), row.names=F)

# generate xtable
print(xtable(ppf), type="html", include.rownames=F)

# plot distributions of total peaks called at each cutoff
library(reshape2)
library(ggplot2)

ppfm <- melt(ppf, id.vars='rpm_cutoff',variable.name='no_peaks') # ppf[m]elted 
g <- ggplot(ppfm, aes(x=as.factor(rpm_cutoff),y=value))
g + geom_boxplot() +
  theme_bw()
ggsave(paste0(path, "a5-peakcut-boxplot.pdf"), width=7, height=5) # export pdf
```
<br>    




