---
title: "T-Treg Gene Expression"
author: "JRD"
date: "March 3, 2016"
output: html_document
---

last modified: `r Sys.Date()`
<br>    

#### README
+ this document contains code and explanation for an analysis of tissue-Treg gene expression signatures
<br>    


#### R Markdown Specifications
Set global options
```{r setoptions, echo=F}
library(knitr)
opts_chunk$set(echo=F, eval=F, warning=F, message=F)
```
<br>    

Github
https://github.com/jdis24/ATAC.git
<br>    


### Pseudo
**divide murine genes into expressed or not in T-Tregs**
+ input = three .gct files from different lab members
+ `Fat-2wkMuscle.gct` - VAT, LN Treg from RB
+ `Gut-PancreasTregs.gct` - Rorc+, Rorc- colonic Treg, no spleens
+ `Dalia (July 23, 2010).gct` - 4d inj muscle, with spleen Treg
+ for each gct, collapse duplicates by median
+ merge gcts together by ProbeSetID
+ draw cv vs mean
+ divide into Treg expr vs non (mean >= 100, cv <= ?)

#### Functions
```{r jrd_functions}
# calculate stats over all numeric columns of df 
row_stats <- function(df) {
  nums <- sapply(df, is.numeric) # extract numeric cols of df
  
  df_nums <- df[,nums] # subset df by numeric cols
  df_nums <- as.data.frame(apply(df_nums, 2, log,2)) # convert expression to log2
  
  stats <- transmute_(df_nums, 
                      Mean = ~rowMeans(df_nums),
                      SD = ~rowSds(as.matrix(df_nums)))
  CV <- stats$SD / stats$Mean
  
  df_other <- df[!nums]     
  cbind(df_other,stats,CV)
}

# calculate stats over any columns of df 
row_stats2 <- function(df,v1,v2,v3) {
  nums <- sapply(df, is.numeric) # extract numeric cols of df
  
  subset <- df[,v1:v3] # subset df by columns
  
  log2subset <- as.data.frame(apply(subset,2,log,2))
    
  means <- transmute_(subset,
                      mean = ~rowMeans(subset),
                      log2mean = ~log(rowMeans(subset),2),
                      log10mean = ~log(rowMeans(subset),10))
  meanlog2 <- transmute_(log2subset,
                         meanlog2 = ~rowMeans(log2subset))
  sd <- transmute_(subset,
                      SD = ~rowSds(as.matrix(subset)))
  sd2 <- transmute_(log2subset,
                      SD = ~rowSds(as.matrix(log2subset)))
  
  cv <- sd$SD / means$mean
  
  cv2 <- sd2$SD / meanlog2$mean
  
  #df_other <- df[!nums]
  cbind(df,means,meanlog2,cv,cv2)
}
```
<br>  


#### Import gcts and collapse duplicate probes by median
```{r gct_collapse, eval=T}
# source cbdm common.R scripts
source("/Volumes/JD_WD2/JRD_CBDMLab/1_JRD_CBDM_PROJECTS/JRD.Scripts/CBDM_wiki_Rscripts/Common.R")

# set path to wd
path <- '/Volumes/JD_WD2/JRD_CBDMLab/JRD_CBDM_public_seqanalyses/Microarray/GCT_given/working'
files <- list.files(path = path, pattern='*.gct', full.names=T)

gctl <- lapply(files, read.gct) # [gctl]ist

fn <- list.files(path = path, pattern='*.gct') # make filenames for each gct
fn <- unlist(strsplit(x = fn, split = '.gct'))
names(gctl) <- fn
# look at colnames
lapply(gctl, names)

# collapse each gct by median expr value
gctl.collapsed <- lapply(gctl, collapseDataset)

# merge by ID, retain IDs present in all
library(dplyr)
# remove ID cols 
gctl.collapsed <- lapply(gctl.collapsed, select, -matches("ID"))
# check that it worked 
lapply(gctl.collapsed, names)
# merge
gctl.joined1 <- inner_join(gctl.collapsed[[1]],gctl.collapsed[[2]], by = "GeneSymbol")
gctl.joined2 <- inner_join(gctl.joined1, gctl.collapsed[[3]], by = "GeneSymbol")
# check names
names(gctl.joined2)

# check expr values in joined match source
gctl.joined2[gctl.joined2$GeneSymbol == "Pparg", colnames(gctl.joined2) =="B614wABDTconv#1"]
gctl[[1]][gctl[[1]]$GeneSymbol == "Pparg", colnames(gctl[[1]]) == "B614wABDTconv#1"]

# save the merged gct
path <- '/Volumes/JD_WD2/JRD_CBDMLab/JRD_CBDM_public_seqanalyses/Microarray/GCT_given/working/'
write.gct(gctl.joined2, filename = paste0(path,'ttreg_gctljoined.gct'))

# remove cols don't want
names(gctl.joined2)
ttreg.gct <- gctl.joined2[,-c(2:43,50:55,62:70)]
names(ttreg.gct)
# need to re-order columns for colonic Tregs
ttreg.gct <- ttreg.gct[,c(1:13,15,17,18,19,14,16,20)]
names(ttreg.gct)

# save the merged, truncated gct
path <- '/Volumes/JD_WD2/JRD_CBDMLab/JRD_CBDM_public_seqanalyses/Microarray/GCT_given/working/'
write.gct(ttreg.gct, filename = paste0(path,'ttreg_mus_spl_vat_cln.gct'))
save(ttreg.gct, file=paste0(path,'ttreg_mus_spl_vat_cln.RDa'))

```
<br> 


#### Plot and filter by cv and expression
```{r cv_mean}
# load file
load(paste0(path,'ttreg_mus_spl_vat_cln.RDa'))
names(ttreg.gct)

## -- compute and apppend stats --------------------------------------------------------------------------
# agct = [a]ppend [gct]
agct <- row_stats2(ttreg.gct, "2","3","4") 
agct <- row_stats2(agct, "5","6","7")
agct <- row_stats2(agct, "8","9","10")
agct <- row_stats2(agct, "11","12","13")
agct <- row_stats2(agct, "14","15","16","17") # changed function to handle 4 cols
agct <- row_stats2(agct, "18","19","20") # changed back to 3

# create column labels for sample stats
stat.labels <- c("_mean","_log2mean","_log10mean","_meanlog2","_cv","_cv2")
sample.names <- strsplit(names(ttreg.gct[-1]), split = "#")
sample.names <- lapply(sample.names, function(x) x[[1]])
sample.names <- unique(unlist(sample.names))

statcol.names <- lapply(sample.names, paste, stat.labels, sep="")
statcol.names <- unlist(statcol.names)

names(agct) <- c(names(agct[1:20]), statcol.names)

# save agct
path <- '/Volumes/JD_WD2/JRD_CBDMLab/JRD_CBDM_public_seqanalyses/Microarray/GCT_given/working/'
save(agct, file=paste0(path,'ttreg_mus_spl_vat_cln_agct.RDa'))


## -- plot expr and cv distributions --------------------------------------------------------------------------
library(reshape2)
library(ggplot2)

# make df of log10 expr for each gene in each sample
l10names <- grep('log10mean$', names(agct))
l10exp <- cbind(agct[1],agct[l10names]) # [l]og[10]exp = new df of log10 expr values
l10expm <- melt(l10exp, id.vars=c('GeneSymbol')) #log10exp[m]elted

# plot
p <- ggplot(l10expm, aes(x=value))
p + geom_histogram(fill="white",color="black", binwidth=0.1) + 
  facet_grid(variable ~.) + 
  geom_vline(xintercept = 2) +
  xlab("log10 expr") + 
  theme_bw()
# save
path <- '/Volumes/JD_WD2/JRD_CBDMLab/JRD_CBDM_public_seqanalyses/Microarray/GCT_given/working/plots/'
ggsave(filename = paste0(path,"log10exp.pdf"), width=8, height=15)


# make df of replicate cv2 (based on log2 expr values) for each gene in each sample
cv2names <- grep("cv2$",names(agct))
cv2cv <- cbind(agct[1],agct[cv2names]) # cv2cv = new df of cv2 values
cv2cvm <- melt(cv2cv, id.vars=c("GeneSymbol"))

# plot
p <- ggplot(cv2cvm, aes(x=GeneSymbol,y=value))
p + geom_point(size=0.1) + 
  facet_grid(variable ~.) + 
  theme_bw() + 
  theme(axis.text.x = element_blank())
# save
path <- '/Volumes/JD_WD2/JRD_CBDMLab/JRD_CBDM_public_seqanalyses/Microarray/GCT_given/working/plots/'
ggsave(filename=paste0(path,"cv2.png"),width=8,height=12)


# make df of log10 expr vs cv2 for each sample
l10cv2 <- cbind(agct[1],agct[l10names],agct[cv2names])
names(l10cv2)

# function for drawing plots 
cv.mean.plot <- function(p,yi) {
  p + geom_point(size=0.2) + 
  geom_hline(yintercept=yi, color="red") + 
  theme_bw() + 
  annotate('text',x=1.5, y=0.3,label=paste('hline at ',yi,sep=""))
}

## if want to plot and save individual cv vs mean plots
p <- ggplot(l10cv2, aes(x=TR.4dINJ.Mu_log10mean,y=TR.4dINJ.Mu_cv2))
cv.mean.plot(p, 0.06)
ggsave(filename=paste0(path,"muscle.cv.mean.pdf"),width=4,height=4)

## if want to plot and save in single image
l10cv2.plot <- l10cv2 # make a copy for melting
samples <- c("Mus.d4","SplwMus.d4","VAT.RB", "LNwVAT.RB","Cln.Rorcneg","Cln.Rorcpos")
names(l10cv2.plot) <- c('GeneSymbol',samples,samples)

df1 <- melt(l10cv2.plot[1:7], id.vars='GeneSymbol', variable.name= 'Sample', value.name ='Mean')
df2 <- melt(l10cv2.plot[c(1,8:13)], id.vars='GeneSymbol', variable.name= 'Sample', value.name ='cv2')
l10cv2.plot <- cbind(df1,df2[3])

yi <- 0.06 # set height of horiz line

# plot across all samples
p <- ggplot(l10cv2.plot, aes(x=Mean,y=cv2))
p + geom_point(size=0.2) + 
  geom_hline(yintercept=0.06, color="red") + 
  theme_bw() + 
  facet_grid(.~Sample) +
  ylim(0,0.4) + 
  ggtitle(paste('log10 expr vs cv(based on log2 expr)', 'hline = ', yi))
  
# save
path <- '/Volumes/JD_WD2/JRD_CBDMLab/JRD_CBDM_public_seqanalyses/Microarray/GCT_given/working/plots/'
ggsave(filename = paste0(path,'log10cv2.pdf'), width=10, height=8)


```













