---
title: "T-Treg Gene Expression"
author: "JRD"
date: "March 3, 2016"
output: html_document
---

last modified: `r Sys.Date()`
<br>    

#### README
+ this document contains code and explanation for an analysis of tissue-Treg gene expression signatures
<br>    


#### R Markdown Specifications
Set global options
```{r setoptions, echo=F}
library(knitr)
opts_chunk$set(echo=F, eval=F, warning=F, message=F)
```
<br>    

Github
https://github.com/jdis24/ATAC.git
<br>    


### Pseudo
**divide murine genes into expressed or not in T-Tregs**
+ input = three .gct files from different lab members
+ `Fat-2wkMuscle.gct` - VAT, LN Treg from RB
+ `Gut-PancreasTregs.gct` - Rorc+, Rorc- colonic Treg, no spleens
+ `Dalia (July 23, 2010).gct` - 4d inj muscle, with spleen Treg
+ for each gct, collapse duplicates by median
+ merge gcts together by ProbeSetID
+ draw cv vs mean
+ divide into Treg expr vs non (mean >= 100, cv <= ?)


#### Import gcts and collapse duplicate probes by median
```{r gct_collapse, eval=T}
# source cbdm common.R scripts
source("/Volumes/JD_WD2/JRD_CBDMLab/1_JRD_CBDM_PROJECTS/JRD.Scripts/CBDM_wiki_Rscripts/Common.R")

# set path to wd
path <- '/Volumes/JD_WD2/JRD_CBDMLab/JRD_CBDM_public_seqanalyses/Microarray/GCT_given/working'
files <- list.files(path = path, pattern='*.gct', full.names=T)

gctl <- lapply(files, read.gct) # [gctl]ist

fn <- list.files(path = path, pattern='*.gct') # make filenames for each gct
fn <- unlist(strsplit(x = fn, split = '.gct'))
names(gctl) <- fn

# collapse each gct by median expr value
gctl.collapsed <- lapply(gctl, collapseDataset)

# merge by ID, retain IDs present in all

test <- 



```
<br> 












