---
title: "TTREG_Microarrays"
author: "JRD"
date: "March 14, 2016"
output: html_document
---

last modified: `r Sys.Date()`
<br>    

#### README
+ this document contains code and explanation for an analysis of tissue-Treg gene expression microarray data
<br>    


#### R Markdown Specifications
Set global options
```{r setoptions, echo=F}
library(knitr)
opts_chunk$set(echo=F, eval=F, warning=F, message=F)
```
<br>    

Github
https://github.com/jdis24/ATAC.git
<br>    


#### Pipeline for pair-wise analysis of each T-Treg with its matching lymphoid Treg (same for Tconv)
+ [following steps done in Excel]
+ trim .gct so that it only contains one set of tissue samples and one set of lymphoid samples
+ put tissue columns before lymphoid
+ [following steps done in R]
+ import .gct file into R
+ collapse duplicate probes by median expression (CBDM code)
+ generate summary stats and plots for each sample: rep mean, rep cv, (dplyr)
+ filter by min expr >100 in >=1 sample and replicate cv < 0.x in all samples
+ calculate fold-change: tissue vs spleen
+ generate FC plots: expr vs expr, p vs fc, fc vs mean (ggplot2)
+ set FC and p-value cutoffs and generate [d]ifferentially [e]xpressed [g]enes (DEG) (Multiplot)
+ annotate DEG (clusterProfiler)
<br>    


#### Functions
```{r functions}
####### ---------------------------- #######
#######       READ GCT FILE          #######
####### ---------------------------- #######
read.gct <- function(datafile="") {
  if(datafile=="") {
    datafile = choose.files(multi=F, filter=c("GCT file","*.gct"))
    if(length(datafile)==0)
      return()
  }	
  x = read.table(datafile, as.is=T, skip=2, header=T, sep="\t", quote="\"", check.names=F, fill=TRUE, comment.char="")
  rownames(x) = as.character(x[,1])
  return(x)
}



###################################
###      COLLAPSE DATASET
###   
### collapse all probe set expression values for 
### a given gene into a single expression value
###
###################################
collapseDataset <- function(gct, collapse.method=median) {
  # Computes a data set with unique gene symbols by collapsing duplicated genes.
  #
  # Args:
  #   gct: A tab-delimited text file of linear gene expression values
  #   collapse.method: the operation used to compute a composite expression value from
  #                    multiple entries of a given gene. Default=median
  #
  # Returns:
  #   a gct file with uniqe entries for each gene in the input gct
  
  # find list of duplicated genes
  duplicates <- unique(gct$GeneSymbol[duplicated(gct$GeneSymbol)])
  
  # collapse gene to a single entry
  out.gct <- matrix(nrow=length(duplicates), ncol=(ncol(gct) - 2))
  for(i in 1:length(duplicates)) {
    gene <- duplicates[i]
    is.gene <- gct$GeneSymbol == gene
    
    out.gct[i,] <- apply((as.matrix(gct[is.gene,][,-c(1:2)])), 2, collapse.method)
    print(paste(i, " Collapsing ", gene))
    
    # replace all values for duplicate genes with values computed by the
    # collapse.method (mean, median, etc...)
    num.cols <- ncol(out.gct)
    num.rows <- sum(is.gene)
    gct[which(is.gene), ][,-c(1,2)] <-  matrix(out.gct[i,], num.rows, num.cols, byrow=T)
  }
  
  # Remove duplicate entries of median/max value for each duplicated gene
  collapsed.gct <- unique(gct[,-1])
  # Remove probeIDs (these are no longer relevant for collapsed genes)
  row.names(collapsed.gct) <- NULL
  return (cbind(ID=1:nrow(collapsed.gct), collapsed.gct))
}


####### ---------------------------- #######
#######           ROW STATS          #######
####### ---------------------------- #######
# calculate stats over all numeric columns of df 
row_stats <- function(df) {
  nums <- sapply(df, is.numeric) # extract numeric cols of df
  
  df_nums <- df[,nums] # subset df by numeric cols
  df_nums <- as.data.frame(apply(df_nums, 2, log,2)) # convert expression to log2
  
  stats <- transmute_(df_nums, 
                      Mean = ~rowMeans(df_nums),
                      SD = ~rowSds(as.matrix(df_nums)))
  CV <- stats$SD / stats$Mean
  
  df_other <- df[!nums]     
  cbind(df_other,stats,CV)
}

```
<br>  



#### Import gct file and collapse duplicate probes
```{r collapse_dups}
## --- set tissue name ------------------------------------------------------------------------------------------
tissue_name <- 'fat'

## --- read gct --------------------------------------------------------------------------------------------------
path <- '/Volumes/JD_WD2/JRD_CBDMLab/JRD_CBDM_public_seqanalyses/Microarray/GCT_given/working/'
file <- paste0(tissue_name,'.gct')

gct <- read.gct(datafile = paste0(path,file)) 

## --- collapse probes -------------------------------------------------------------------------------------------
gct_coll <- collapseDataset(gct = gct)

# remove first two columns
gct_coll <- gct_coll[-1]

## --- write csv -------------------------------------------------------------------------------------------------
write.csv(gct_coll, file = paste0(path,tissue_name,'.collapsed','.csv'), row.names = F)

```
<br>  


#### Sample summary stats and plots
```{r summ_stats}
# divide into tissue and lymphoid df
# add sampleID
# combine into list
# apply stat function
# melt the list
# generate plots, facet on SampleID
## --- import file ------------------------------------------------------------------------------------------------
path <- '/Volumes/JD_WD2/JRD_CBDMLab/JRD_CBDM_public_seqanalyses/Microarray/GCT_given/working/'
tissue_name <- 'fat'
dat <- read.csv(file = paste0(path,tissue_name,'.collapsed','.csv'))


## --- divide samples ---------------------------------------------------------------------------------------------
tdat <- data.frame(dat[1:3])
ldat <- data.frame(dat[c(1,4:5)])

## --- add sample ID  --------------------------------------------------------------------------------------------
tname <- 'VAT'
lname <- 'LN'

tdat <- cbind(tdat, name = tname)
ldat <- cbind(ldat, name = lname)

## --- combine into list ------------------------------------------------------------------------------------------
sl <- list(tdat, ldat)
names(sl) <- c(tname, lname)

## --- generate replicate stats -----------------------------------------------------------------------------------
library(dplyr)
library(matrixStats)
sl.stats <- lapply(sl, row_stats)

## --- plot ------------------------------------------------------------------------------------------------------
library(reshape2)
library(ggplot2)
library(scales)
sl.stats.m <- melt(sl.stats, id.vars = c('GeneSymbol','name','Mean','SD','CV'))
sl.stats.m <- sl.stats.m[-6]

# plot cv vs log2 mean
p <- ggplot(sl.stats.m, aes(x=Mean, y=CV))
# set cv line to draw
cut <- 0.06
p + 
  geom_point(size = 0.6, alpha = 0.2) +
  facet_wrap(~ name) + 
  scale_x_continuous(breaks = pretty_breaks(), limits = 
                       c(round(min(sl.stats.m$Mean)),round(max(sl.stats.m$Mean)))) +
  xlab("Mean Expression (log2)") + 
  geom_hline(yintercept = 0.06, color = "red") +
  ggtitle(paste('CV vs log2 mean ','cv line = ',cut,sep="")) +
  theme_bw()

# save
path <- '/Volumes/JD_WD2/JRD_CBDMLab/JRD_CBDM_public_seqanalyses/Microarray/GCT_given/working/'
ggsave(file = paste0(path,tissue_name,'.repcv_vs_mean','.pdf'), width = 8, height = 4)


##  
```
<br>  


#### Fold-change plots
```{r fcplots}
mean.cv <- data.frame(GeneSymbol = sl.stats[[1]]$GeneSymbol,
                   Tissue_mean = sl.stats[[1]]$Mean, 
                   Lymphoid_mean = sl.stats[[2]]$Mean,
                   Tissue_cv = sl.stats[[1]]$CV,
                   Lymphoid_cv = sl.stats[[2]]$CV)

# set cutoffs
mean <- 100
cv <- 0.06

# filter by mean

mean.fil <- 

```
<br>


#### Set cutoffs and generate DEG
```{r DEG}
```
<br> 











